Palmer amaranth (\emph{Amaranthus palmeri}) adaptation to US Midwest agroecosystems

### 2021-09-16 Manuscript/Manuscript.tex
as

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is currently
ranked as one of the most economically detrimental weed species to
cropping systems in the United States (Van Wychen, 2020). Unmanaged
Palmer amaranth plants compete for water, light, and nutrients, which
can drastically impact crop yields (Berger et al., 2015). For example,
Palmer amaranth has been documented to reduce up to 91\%, 68\%, and 54\%
corn (Massinga et al., 2001), soybean (Klingaman and Oliver, 1994), and
cotton (Morgan et al., 2001) yields, respectively. Moreover, Palmer
amaranth has shown a remarkable capacity to evolve resistance to
herbicides. To date, Palmer amaranth has evolved resistance to eight
herbicide sites of action (Heap, 2021), increasing the weed management
complexity (Lindsay et al., 2017). Thus, Palmer amaranth poses an
economical and environmental risk to sustainable agriculture.

Palmer amaranth is a fast growing summer annual forb indigenous to the
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with plant
biology. Farm mechanization, adoption of conservation agriculture (e.g.,
no-till), and intensive use of herbicides for weed management are the
main human-mediated selection of Palmer amaranth into cropping systems
(Ward et al., 2013). On the other hand, Palmer amaranth is a prolific
seed producer with a C4 photosynthetic apparatus (Wang et al., 1992).
With a dioecy nature, Palmer amaranth male and female plants are
obligate outcrosser species, increasing the chances of exchanging
adaptive traits among plants (Jhala et al., 2021; Oliveira et al.,
2018). Also, Palmer amaranth small seeds (e.g, 1 mm) tend to thrive in
no-tillage systems (Price et al., 2011), and spread across locations
through farm equipment (Sauer, 1972), manure (Hartzler and Anderson,
2016), wildlife (Farmer et al., 2017), etc. The dispersal capacity of
Palmer amaranth makes the species one of the most successful cases of
weed adaptation to current cropping systems.

Light and temperature are the main environmental requirements for Palmer
amaranth growth and development (Jha et al., 2010). Palmer amaranth is
reported with an extended germination period (Ward et al., 2013).
Germination of Palmer amaranth was triggered by 18 C soil temperature at
5 cm depth (Keeley et al., 1987), and optimal germination and biomass
production occurred at 35/30 C day and night temperatures (Guo and
Al-Khatib, 2003). In addition, Palmer amaranth establishment is
human-mediated by tillage timings and preemergence-applied herbicides
(Chahal et al., 2021), which can result in weed germination shifts
(Sbatella and Wilson, 2010). Soil water content has not shown to limit
Palmer amaranth fitness. Under continuous water stress, Palmer amaranth
survived and produced at least 14000 seeds plant\textsuperscript{-1}
(Chahal et al., 2018). Seeds from Palmer amaranth growing with limited
water conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). Growing conditions and management practices
also influence Palmer amaranth sex dimorphism and flowering pattern
(Korres et al., 2017; Rumpa et al., 2019). Therefore, Palmer amaranth
has shown plasticity to evolve and fast adapt under the current
agroecosystem conditions. Further scenarios show that global temperature
warming can impact agriculture, and promote niches for Palmer amaranth
invasion/adaptation into new environments. Currently, it is estimated
that the greatest climatic risk of Palmer amaranth establishment are
agronomic crops in Australia and Sub-Sahara Africa (Kistner and
Hatfield, 2018). Temperature is a key factor limiting Palmer amaranth
expansion to cooler geographies (Briscoe Runquist et al., 2019);
however, under future climate change Palmer amaranth is likely to expand
northward into Canada and Northern Europe (Kistner and Hatfield, 2018;
Briscoe Runquist et al., 2019).

@@

\caption{\label{tab:unnamed-chunk-2}Field study attributes from Arlington, WI, Clay Center, NE, Grant, NE, Lincoln, NE and Macomb, IL}
\centering
\resizebox{\linewidth}{!}{
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lllllll}
\toprule
Attributes &  & Arlington, WI & Clay Center, NE & Grant, NE & Lincoln, NE & Macomb, IL\\
\midrule
Bareground &  &  &  &  &  & \\

 & Weed control & glyphosate\textsuperscript{a} / \emph{S}-metolachor\textsuperscript{b} & saflufenacil + imazethapyr + pyroxasulfone\textsuperscript{c} &  &  & \\

@@

 & Weed control & glyphosate / \emph{S}-metolachor & \emph{S}-metolachlor + trazine + mesotrione + bicyclopyrone\textsuperscript{d} &  &  & \\

 & Stage at 1\textsuperscript{st} cohort & V2-3 &  &  &  & \\

 & Stage at 2\textsuperscript{nd} cohort & V6-7 &  &  &  & \\

@@

 & Weed control & glyphosate / \emph{S}-metolachor & saflufenacil + imazethapyr + pyroxasulfone &  &  & \\

 & Stage at 1\textsuperscript{st} cohort & V1-2 &  &  &  & \\

 & Stage at 2\textsuperscript{nd} cohort & V5-6 &  &  &  & \\

@@

\multirow{-4}{*}{\raggedright\arraybackslash } & Organic matter (\%) & 3.5 & 3 &  &  & \\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a} glyphoste, 840 g ae ha \textsuperscript{b} \emph{S}--metolachor, 1324 g ai ha; \textsuperscript{c} saflufenacil + imazethapyr + pyroxasulfone, 215 g ai ha; \textsuperscript{d} \emph{S}-metolachlor + trazine + mesotrione, + bicyclopyrone, 2409 g ai ha}\\
\end{tabular}}
\end{table}

@@

The statistical analyses were performed using R statistical software
version 4.0.1 (Team, 2021). Data analyses are stored at Zenodo
(Oliveira, 2021).

@@

An ecological approach to reduce Palmer amaranth seed production is
understanding plant biology, including flowering pattern. Our study
suggests that Palmer amaranth flowering was slightly influenced by crops
and cohort timings. Palmer amaranth growing in bareground and corn
resulted in the overall shortest and longest flowering window,
respectively. The shortest flowering window of second cohorts of Palmer
amaranth growing in bareground highlighted the impact of photoperiod on
flowering. When growing in soybean, Palmer amaranth flowering window was
similar to bareground at first cohort but similar to corn at second
cohort timing. Plant flowering initiation is complex and depends on
biological and ecological factors (Lang, 1965). We hypothesize that when
growing in high competition (e.g., second cohort), Palmer amaranth
plants tend to initiate flowering early, as well as having an extended
flowering window. A study has shown that Palmer amaranth initiated
flowering two weeks prior to the native waterhemp in Iowa (Baker, 2021).
Early flower initiation could also be a plant strategy when growing in
stress conditions. For example, when growing under water stress, early
flowering in Palmer amaranth resulted in a mismatch between female and
male plants by seven days (Mesgaran et al., 2021). A mismatch in Palmer
amaranth male and female flowering period can minimize plant outcross,
reducing plant seed production and/or exchange of resistant alleles
(Jhala et al., 2021). Sex dimorphism manipulation is considered a
potential ecological pest control (McFarlane et al., 2018; Schliekelman
et al., 2005).

@@

Our study demonstrated the Palmer amaranth plasticity to grow and
develop into arable land of US Midwest. It is likely that Palmer
amaranth range will continue to expanding into new geographies. The
migration of Palmer amaranth into the US Midwest will reshape the
landscape as waterhemp and Palmer amaranth will share the same habitat.
The presence of Palmer amaranth and waterhemp will increase the weed
management complexity. Therefore, preventive management is a priority to
minimizing Palmer amaranth dispersal. Reactive management should focus
on early-season management programs, which would have a large negative
effect on Palmer amaranth growth and development. Long-term tactics that
promote early-season crop advantage against Palmer amaranth, including
diversity of crops in rotation, early/late crop planting, plant width,
and crop residue (e.g.~cover crops) would minimize the negative impact
of Palmer amaranth to cropping systems. The aggressiveness and
differential Palmer amaranth adaptation to agroecosystem is striking and
require national efforts to minimize the species impact on
sustainability and profitability of cropping systems.

@@

>>>>

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is currently
ranked as one of the most economically detrimental weed species to
cropping systems in the United States (Van Wychen, 2020). Unmanaged
Palmer amaranth plants compete for water, light, and nutrients, which
can drastically impact crop yields (Berger et al., 2015). For example,
Palmer amaranth has been documented to reduce up to 91\%, 68\%, and 54\%
corn (Massinga et al., 2001), soybean (Klingaman and Oliver, 1994), and
cotton (Morgan et al., 2001) yields, respectively. Moreover, Palmer
amaranth has shown a remarkable capacity to evolve resistance to
herbicides. To date, Palmer amaranth has evolved resistance to eight
herbicide sites of action (Heap, 2021), increasing the weed management
complexity (Lindsay et al., 2017). Thus, Palmer amaranth poses an
economical and ecological risk to sustainable agriculture.

Palmer amaranth is a fast growing summer annual forb indigenous to the
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with plant
biology. Farm mechanization, adoption of conservation agriculture (e.g.,
no-till), and intensive use of herbicides for weed management are the
main human-mediated selection of Palmer amaranth into cropping systems
(Ward et al., 2013). On the other hand, Palmer amaranth is a prolific
seed producer with a C4 photosynthetic apparatus (Wang et al., 1992).
With a dioecy nature, Palmer amaranth male and female plants are
obligate outcrosser species, increasing the chances of exchanging
adaptive traits among plants (Jhala et al., 2021; Oliveira et al.,
2018). Also, Palmer amaranth small seeds (e.g, 1 mm) tend to thrive in
no-tillage systems (Price et al., 2011), and spread across locations
through farm equipment (Sauer, 1972), seed mixes (Hartzler and Anderson,
2016), wildlife (Farmer et al., 2017), etc. The dispersal capacity of
Palmer amaranth makes the species one of the most successful cases of
weed adaptation to current cropping systems.

Light and temperature are the main environmental requirements for Palmer
amaranth growth and development (Jha et al., 2010). Palmer amaranth is
reported with an extended germination period (Ward et al., 2013).
Germination of Palmer amaranth was triggered by 18 C soil temperature at
5 cm depth (Keeley et al., 1987), and optimal germination and biomass
production occurred at 35/30 C day and night temperatures (Guo and
Al-Khatib, 2003). In addition, Palmer amaranth establishment is
human-mediated by tillage timings and preemergence-applied herbicides
(Chahal et al., 2021), which can result in weed germination shifts
(Sbatella and Wilson, 2010). Also, Palmer amaranth has shown to grow and
develop under reduced soil water content. In continuous water stress,
Palmer amaranth survived and produced at least 14000 seeds
plant\textsuperscript{-1} (Chahal et al., 2018). Seeds from Palmer
amaranth growing with limited water conditions were heavier, less
dormant, and prompt for germination (Matzrafi et al., 2021). Growing
conditions and management practices also influence Palmer amaranth sex
dimorphism and flowering pattern (Korres et al., 2017; Rumpa et al.,
2019). Therefore, Palmer amaranth has shown plasticity to evolve and
fast adapt under the current agroecosystem conditions. Further scenarios
show that global temperature warming can impact agriculture, and promote
niches for Palmer amaranth invasion/adaptation into new environments.
Currently, it is estimated that the greatest climatic risk of Palmer
amaranth establishment are agronomic crops in Australia and Sub-Sahara
Africa (Kistner and Hatfield, 2018). Temperature is a key factor
limiting Palmer amaranth expansion to cooler geographies (Briscoe
Runquist et al., 2019); however, under future climate change Palmer
amaranth is likely to expand northward into Canada and Northern Europe
(Kistner and Hatfield, 2018; Briscoe Runquist et al., 2019).

@@

\caption{\label{tab:unnamed-chunk-2}Field study attributes }
\centering
\resizebox{\linewidth}{!}{
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lllllll}
\toprule
Attributes &  & Arlington, WI & Clay Center, NE & Grant, NE & Lincoln, NE & Macomb, IL\\
\midrule
Bareground &  &  &  &  &  & \\

 & Weed control & glyphosate & saflufenacil + imazethapyr + pyroxasulfone &  &  & \\

@@

 & Weed control & glyphosate / S-metolachor & S-metolachlor + trazine + mesotrione, + bicyclopyrone &  &  & \\

 & Stage at 1 cohort & V2-3 &  &  &  & \\

 & Stage at 2 cohort & V6-7 &  &  &  & \\

@@

 & Weed control & glyphosate / S-metolachor & saflufenacil + imazethapyr + pyroxasulfone &  &  & \\

 & Stage at 1 cohort & V1-2 &  &  &  & \\

 & Stage at 2 cohort & V5-6 &  &  &  & \\

@@

\multirow{-4}{*}{\raggedright\arraybackslash } & Organic matter (\%) & 3.5 & 3 &  &  & \\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a} glyphoste, 840 g ae ha \textsuperscript{b} S-metolachor, 1324 g ai ha; \textsuperscript{c} S-metolachlor + trazine + mesotrione, + bicyclopyrone, 2409 g ai ha; \textsuperscript{d} saflufenacil + imazethapyr + pyroxasulfone, 215 g ai ha}\\
\end{tabular}}
\end{table}

@@

The statistical analyses were performed using R statistical software
version 4.0.1 (Team, 2021).

@@

An ecological approach to reduce Palmer amaranth seed production is
understanding plant biology, including flowering pattern. Our study
suggests that Palmer amaranth flowering was slightly influenced by crops
and cohort timings. Palmer amaranth growing in bareground and corn
resulted in the overall shortest and longest flowering window,
respectively. The shortest flowering window of second cohorts of Palmer
amaranth growing in bareground highlighted the impact of photoperiod on
flowering. When growing in soybean, Palmer amaranth flowering window was
similar to bareground at first cohort but similar to corn at second
cohort timing. Plant flowering initiation is complex and depends on the
complex interaction between genetic makeup and the environmental
conditions (Lang, 1965). We hypothesize that when growing in high
competition (e.g., second cohort), Palmer amaranth plants tend to
initiate flowering early, as well as having an extended flowering
window. A study has shown that Palmer amaranth initiated flowering two
weeks prior to the native waterhemp in Iowa (Baker, 2021). Early flower
initiation could also be a plant strategy when growing in stress
conditions. For example, when growing under water stress, early
flowering in Palmer amaranth resulted in a mismatch between female and
male plants by seven days (Mesgaran et al., 2021). A mismatch in Palmer
amaranth male and female flowering period can minimize plant outcross,
reducing plant seed production and/or exchange of resistant alleles
(Jhala et al., 2021). Sex dimorphism manipulation is considered a
potential ecological pest control (McFarlane et al., 2018; Schliekelman
et al., 2005).

@@

Our study demonstrated the Palmer amaranth plasticity to grow and
develop into arable land of US Midwest. It is likely that Palmer
amaranth range will continue to expanding into new geographies. The
migration of Palmer amaranth into the US Midwest will reshape the
landscape as waterhemp and Palmer amaranth will share the same habitat.
The presence of Palmer amaranth and waterhemp will increase the weed
management complexity. Therefore, preventive management is a priority to
minimizing Palmer amaranth dispersal. Reactive management should focus
on early-season management programs, which would have a large negative
effect on Palmer amaranth growth and development. Long-term tactics that
promote early-season crop advantage against Palmer amaranth, including
diversity of crops in rotation, early/late crop planting, plant width,
and crop residue (e.g.~cover crops) would minimize the negative impact
of Palmer amaranth to cropping systems. The aggressiveness and
differential Palmer amaranth adaptation to agroecosystem is striking and
require regional collaboration to minimize the species impact on
sustainability and profitability of cropping systems.

@@

\hypertarget{data-accessibility}{%
\section*{Data accessibility}\label{data-accessibility}}
\addcontentsline{toc}{section}{Data accessibility}

The data and scripts used to analyse the data presented in this work can
be found at Zenodo (Oliveira, 2021).
